---
title: "NFL game predictor"
subtitle: Keegan Prunty
date: "2025-11-12"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Objective

The goal of this project is to predict whether a home team will win an NFL 
regular season game using cumulative information on the team's season. This was
done so that it can predict games that have not occured yet based on pre-kickoff
information. We will create a data set based on NFL season to date team stats 
and fit a Logistic Regression, along with feature selection for the best model.

# Data Background

nflreadr::load_schedules(2024) for game-level results 
(teams, scores, week, game_id).

nflreadr::load_pbp(2024) for play-by-play data to compute drive-based stats, 
third-down rates, turnovers, and EPA.

(For this specific model we will be using 2024 regular season data)

# Data Manipulation and Cleaning

We build a table with one row per team per game (team_game) containing:
1. Points for / against
2. Win indicator
3. Turnover margin (takeaways - giveaways)
4. Drive scoring rate (scoring drives / total drives)
5. Time of possession (seconds)
6. Third-down success rate
7. Offensive EPA per play

These chosen stats are often key indicators of team success to an avid NFL fan. 

```{r}
# packages
library(dplyr)
library(nflreadr)
library(slider)
library(tidyr)

season <- 2024

# ---------- 1) Load schedules & pbp ----------
sched <- load_schedules(season) %>%
  filter(game_type == "REG") %>%
  select(game_id, season, week, home_team, away_team, home_score, away_score)

pbp <- load_pbp(season) %>%
  # keep regular season only
  inner_join(sched %>% select(game_id), by = "game_id")

# ---------- 2) Build team-game base (points for/against) ----------
team_game <- bind_rows(
  sched %>% transmute(season, week, game_id,
                      team = home_team, opp_team = away_team,
                      pts_for = home_score, pts_against = away_score, 
                      home = 1L),
  sched %>% transmute(season, week, game_id,
                      team = away_team, opp_team = home_team,
                      pts_for = away_score, pts_against = home_score, home = 0L)
) %>%
  arrange(season, team, week) %>%
  group_by(season, team) %>%
  mutate(
    win = as.integer(pts_for > pts_against),
    cum_wins = lag(cumsum(win)),
    games_played = lag(row_number() - 1L),
    win_pct = ifelse(games_played > 0, cum_wins / games_played, NA_real_),
    last3_win_pct = lag(slide_dbl(win, mean, .before = 3, .complete = FALSE))
  ) %>%
  ungroup()

# ---------- 3) Turnovers: giveaways (offense) & takeaways (defense) ----------
turnovers_off <- pbp %>%
  filter(!is.na(posteam)) %>%
  group_by(game_id, team = posteam) %>%
  summarise(
    giveaways = sum(replace_na(interception, 0)) + 
      sum(replace_na(fumble_lost, 0)),
    .groups = "drop"
  )

# defensive takeaways = opponent’s giveaways (map via opponent)
turnovers_def <- turnovers_off %>%
  rename(opp_team = team, takeaways = giveaways)

team_game <- team_game %>%
  left_join(turnovers_off, by = c("game_id","team")) %>%
  left_join(turnovers_def, by = c("game_id","opp_team")) %>%
  mutate(
    giveaways = replace_na(giveaways, 0),
    takeaways = replace_na(takeaways, 0),
    to_margin = takeaways - giveaways
  ) %>%
  group_by(season, team) %>%
  mutate(
    cum_to_margin = lag(cumsum(to_margin)),       # cumulative turnover margin
    to_margin_per_game = ifelse(games_played > 0, cum_to_margin / games_played, NA_real_)
  ) %>% ungroup()

# ---------- 4) Scoring % (drive-based) ----------
# Build drive-level table (offensive drives only) from pbp
# helper: parse "mm:ss" to seconds (NA-safe)
parse_mmss <- function(x) {
  x <- ifelse(is.na(x) | x == "", NA, x)
  parts <- strsplit(x, ":", fixed = TRUE)
  as.numeric(vapply(parts, function(p)
    if (length(p) == 2) as.integer(p[1]) * 60 + 
      as.integer(p[2]) else NA_real_, 1.0))
}

drives <- pbp %>%
  filter(!is.na(posteam), !is.na(drive)) %>% #keep offensive plays with id
  group_by(game_id, team = posteam, drive) %>%
  summarise(
    # scoring: any play on this drive marked as ended-with-score
    drive_scored = as.integer(any(coalesce(drive_ended_with_score, 0) == 1)),
    # take the last non-NA TOP string for the drive (it's "mm:ss")
    drive_top_mmss = dplyr::last(na.omit(drive_time_of_possession)),
    .groups = "drop_last"
  ) %>%
  # parse mm:ss to numeric seconds
  mutate(drive_seconds = parse_mmss(drive_top_mmss)) %>%
  # collapse to team-game
  group_by(game_id, team) %>%
  summarise(
    drives_total  = n(),
    drives_scored = sum(drive_scored, na.rm = TRUE),
    top_seconds   = sum(coalesce(drive_seconds, 0), na.rm = TRUE),
    .groups = "drop"
  )

team_game <- team_game %>%
  left_join(drives, by = c("game_id","team")) %>%
  mutate(
    drives_total = replace_na(drives_total, 0),
    drives_scored = replace_na(drives_scored, 0),
    top_seconds = replace_na(top_seconds, 0),
    score_pct_game = ifelse(drives_total > 0, drives_scored / drives_total, 
                            NA_real_)
  ) %>%
  group_by(season, team) %>%
  mutate(
    score_pct_accum = lag(cummean(score_pct_game)),          # cum scoring %
    top_per_game = lag(cumsum(top_seconds) / (games_played)) # cumavg time
  ) %>% ungroup()

# ---------- 5) 3rd-down success rate (offense) ----------

# --- Offensive 3rd-down per team-game ---
third_off <- pbp %>%
  filter(!is.na(posteam), down == 3) %>%
  group_by(game_id, team = posteam) %>%
  summarise(
    third_conv = sum(as.integer(coalesce(third_down_converted, 0))),
    third_fail = sum(as.integer(coalesce(third_down_failed, 0))),
    third_att  = third_conv + third_fail,
    third_sr_game = ifelse(third_att > 0, third_conv / third_att, NA_real_),
    .groups = "drop"
  )

# --- Defensive 3rd-down per team-game (allowed) ---
third_def <- pbp %>%
  filter(!is.na(defteam), down == 3) %>%
  group_by(game_id, team = defteam) %>%
  summarise(
    third_conv_allowed = sum(as.integer(coalesce(third_down_converted, 0))),
    third_fail_forced  = sum(as.integer(coalesce(third_down_failed, 0))),
    third_att_def      = third_conv_allowed + third_fail_forced,
    third_sr_allowed_game = ifelse(third_att_def > 0, third_conv_allowed / 
                                     third_att_def, NA_real_),
    .groups = "drop"
  )

# Join into your team_game table and make cumulative (season-to-date) versions
team_game <- team_game %>%
  left_join(third_off, by = c("game_id","team")) %>%
  left_join(third_def, by = c("game_id","team")) %>%
  group_by(season, team) %>%
  mutate(
    third_sr_accum_off = lag(cummean(third_sr_game)),          # offense S2D
    third_sr_accum_def = lag(cummean(third_sr_allowed_game)),  # defense S2D
    third_att_cum_off  = lag(cumsum(coalesce(third_att, 0))),  #count
    third_att_cum_def  = lag(cumsum(coalesce(third_att_def, 0)))
  ) %>%
  ungroup()

# ---------- 6) EPA (offense & defense), per game and cumulative ----------
epa_game <- pbp %>%
  filter(!is.na(posteam)) %>%
  group_by(game_id, team = posteam) %>%
  summarise(epa_off_play = mean(epa, na.rm = TRUE), .groups = "drop")

epa_allowed_game <- pbp %>%
  filter(!is.na(defteam)) %>%
  group_by(game_id, team = defteam) %>%
  summarise(epa_def_allowed = mean(epa, na.rm = TRUE), .groups = "drop")

team_game <- team_game %>%
  left_join(epa_game, by = c("game_id","team")) %>%
  left_join(epa_allowed_game, by = c("game_id","team")) %>%
  group_by(season, team) %>%
  mutate(
    epa_off_cum = lag(cummean(epa_off_play)),        # cum offensive EPA/play
    epa_def_cum = lag(cummean(epa_def_allowed))      # cum defensive EPA/play 
  ) %>% ungroup()

# ---------- 7) Opponent pregame win% and cumulative SOS ----------
opp_wp <- team_game %>%
  select(season, week, team, win_pct) %>%
  rename(opp_team = team, opp_win_pct_before = win_pct)

team_game <- team_game %>%
  left_join(opp_wp, by = c("season","week","opp_team")) %>%
  group_by(season, team) %>%
  mutate(
    sos_accum = lag(cummean(opp_win_pct_before))     # difficulty of schedule
  ) %>% ungroup()

# ---------- add cumulative (pregame) metrics before feature export ----------
team_game <- team_game %>%
  mutate(week = as.integer(week)) %>%
  arrange(season, team, week) %>%
  group_by(season, team) %>%
  mutate(
    # games / wins before current game
    gp_before   = lag(row_number(), default = 0L),
    wins_cum    = cumsum(win),
    wins_before = lag(wins_cum, default = 0L),
    win_pct     = ifelse(gp_before > 0, wins_before / gp_before, NA_real_),

    # cumulative points for/against (pregame PPG)
    pts_for_cum_before     = lag(cumsum(pts_for), default = 0),
    pts_against_cum_before = lag(cumsum(pts_against), default = 0),
    ppg_for_cum            = ifelse(gp_before > 0, pts_for_cum_before / 
                                      gp_before, NA_real_),
    ppg_against_cum        = ifelse(gp_before > 0, pts_against_cum_before / 
                                      gp_before, NA_real_),

    # cumulative turnover margin
    to_margin_cum_before   = lag(cumsum(to_margin), default = 0),
    to_margin_per_game     = ifelse(gp_before > 0, to_margin_cum_before / 
                                      gp_before, NA_real_),

    # cumulative scoring %, TOP, third downs, EPA, etc.
    drives_scored_before   = lag(cumsum(coalesce(drives_scored, 0)), 
                                 default = 0),
    drives_total_before    = lag(cumsum(coalesce(drives_total, 0)),  
                                 default = 0),
    score_pct_accum        = ifelse(drives_total_before > 0, 
                                    drives_scored_before / drives_total_before, 
                                    NA_real_),

    top_seconds_before     = lag(cumsum(coalesce(top_seconds, 0)), default = 0),
    top_per_game           = ifelse(gp_before > 0, top_seconds_before / 
                                      gp_before, NA_real_),

    third_conv_cum_off     = lag(cumsum(coalesce(third_conv, 0)), default = 0),
    third_att_cum_off      = lag(cumsum(coalesce(third_att, 0)),  default = 0),
    third_sr_accum_off     = ifelse(third_att_cum_off > 0, third_conv_cum_off / 
                                      third_att_cum_off, NA_real_),

    third_conv_cum_def     = lag(cumsum(coalesce(third_conv_allowed, 0)), 
                                 default = 0),
    third_att_cum_def      = lag(cumsum(coalesce(third_att_def, 0)),      
                                 default = 0),
    third_sr_accum_def     = ifelse(third_att_cum_def > 0, third_conv_cum_def / 
                                      third_att_cum_def, NA_real_),

    epa_off_sum_before     = lag(cumsum(coalesce(epa_off_play, NA_real_)), 
                                 default = NA),
    epa_off_cnt_before     = lag(cumsum(!is.na(epa_off_play)), default = 0),
    epa_off_cum            = ifelse(epa_off_cnt_before > 0, epa_off_sum_before /
                                      epa_off_cnt_before, NA_real_),

    epa_def_sum_before     = lag(cumsum(coalesce(epa_def_allowed, NA_real_)), 
                                 default = NA),
    epa_def_cnt_before     = lag(cumsum(!is.na(epa_def_allowed)), default = 0),
    epa_def_cum            = ifelse(epa_def_cnt_before > 0, epa_def_sum_before / 
                                      epa_def_cnt_before, NA_real_)
  ) %>% 
  ungroup()

# ---------- fix ppg ----------
team_game <- team_game %>%
  mutate(week = as.integer(week)) %>%
  arrange(season, team, week) %>%
  group_by(season, team) %>%
  mutate(
    gp_before        = lag(row_number(), default = 0L), #games before
    wins_cum         = cumsum(win),
    wins_before      = lag(wins_cum, default = 0L),
    win_pct          = ifelse(gp_before > 0, wins_before / gp_before, NA_real_)
  ) %>% ungroup()

team_game <- team_game %>%
  arrange(season, team, week) %>%
  group_by(season, team) %>%
  mutate(
    # games played before this one
    gp_before = lag(row_number(), default = 0L),

    # cumulative totals before this week
    pts_for_before     = lag(cumsum(pts_for), default = 0),
    pts_against_before = lag(cumsum(pts_against), default = 0),

    # pregame average (PPG)
    ppg_for_cum     = ifelse(gp_before > 0, pts_for_before / gp_before, 
                             NA_real_),
    ppg_against_cum = ifelse(gp_before > 0, pts_against_before / gp_before, 
                             NA_real_)
  ) %>%
  ungroup()



# ---------- 9) Final: select what you need ----------
team_game_features <- team_game %>%
  transmute(
    season, week, game_id, team, opp_team, home,
    pts_for, pts_against, ppg_for_cum, ppg_against_cum, win, win_pct, 
    last3_win_pct,
    giveaways, takeaways, to_margin, cum_to_margin, to_margin_per_game,
    drives_total, drives_scored,
    score_pct_game = coalesce(score_pct_game, NA_real_),
    score_pct_accum = coalesce(score_pct_accum, NA_real_),
    top_seconds = coalesce(top_seconds, NA_real_),
    top_per_game = coalesce(top_per_game, NA_real_),

    third_sr_game_off  = coalesce(third_sr_game, NA_real_),
    third_sr_accum_off = coalesce(third_sr_accum_off, NA_real_),
    third_sr_game_def  = coalesce(third_sr_allowed_game, NA_real_),
    third_sr_accum_def = coalesce(third_sr_accum_def, NA_real_),

    epa_off_play  = coalesce(epa_off_play, NA_real_),
    epa_off_cum   = coalesce(epa_off_cum, NA_real_),
    epa_def_allowed = coalesce(epa_def_allowed, NA_real_),
    epa_def_cum   = coalesce(epa_def_cum, NA_real_),

    opp_win_pct_before = coalesce(opp_win_pct_before, NA_real_),
    sos_accum = coalesce(sos_accum, NA_real_)
  )
```

This is our model approach using these final mathcup diferentials.

```{r}
# Create home/away feature tables for joining onto schedule
home_feats <- team_game_features %>%
  transmute(
    game_id,
    home_team = team,
    home_ppg_for_cum       = ppg_for_cum,
    home_ppg_against_cum   = ppg_against_cum,
    home_epa_off_cum       = epa_off_cum,
    home_to_margin_per_game= to_margin_per_game,
    home_score_pct_accum   = score_pct_accum,
    home_third_sr_accum_off= third_sr_accum_off,
    home_top_per_game      = top_per_game,
    home_win_pct           = win_pct,
    home_last3_win_pct     = last3_win_pct
  )

away_feats <- team_game_features %>%
  transmute(
    game_id,
    away_team = team,
    away_ppg_for_cum       = ppg_for_cum,
    away_ppg_against_cum   = ppg_against_cum,
    away_epa_off_cum       = epa_off_cum,
    away_to_margin_per_game= to_margin_per_game,
    away_score_pct_accum   = score_pct_accum,
    away_third_sr_accum_off= third_sr_accum_off,
    away_top_per_game      = top_per_game,
    away_win_pct           = win_pct,
    away_last3_win_pct     = last3_win_pct
  )

matchups <- sched %>%
  left_join(home_feats, by = c("game_id","home_team")) %>%
  left_join(away_feats, by = c("game_id","away_team")) %>%
  mutate(
    # Target variable: 1 if home team won
    home_win = as.integer(home_score > away_score),

    # Feature differentials (home - away)
    d_ppg_for_cum     = home_ppg_for_cum     - away_ppg_for_cum,
    d_ppg_against_cum = home_ppg_against_cum - away_ppg_against_cum,
    d_epa_off_cum     = home_epa_off_cum     - away_epa_off_cum,
    d_to_margin_pg    = home_to_margin_per_game - away_to_margin_per_game,
    d_score_pct       = home_score_pct_accum - away_score_pct_accum,
    d_third_sr_off    = home_third_sr_accum_off - away_third_sr_accum_off,
    d_top_per_game    = home_top_per_game    - away_top_per_game,
    d_win_pct         = home_win_pct         - away_win_pct,
    d_last3_win_pct   = home_last3_win_pct   - away_last3_win_pct
  )
```

```{r}
#picking the features I want and dropping NA for early games with missing stats
model_data <- matchups %>%
  filter(!is.na(d_ppg_for_cum)) %>%
  select(
    home_win,
    d_ppg_for_cum,
    d_ppg_against_cum,
    d_epa_off_cum,
    d_to_margin_pg,
    d_score_pct,
    d_third_sr_off,
    d_top_per_game,
    d_win_pct,
    d_last3_win_pct
  )
```

# Logistic Regression

Let's create our model and take a look at our features.

```{r}
#splitting data
set.seed(42)
train_index <- sample(seq_len(nrow(model_data)), size = 0.8 * nrow(model_data))
train_data <- model_data[train_index, ]
test_data  <- model_data[-train_index, ]


#fitting model logistoc regression
model_logit <- glm(home_win ~ ., data = train_data, family = binomial)
summary(model_logit)
exp(coef(model_logit))
test_data$pred_prob <- predict(model_logit, newdata = test_data, 
                               type = "response")
test_data$pred_win  <- ifelse(test_data$pred_prob > 0.5, 1, 0)
```

As you can see non of the features seem to be significant above and we can run 
into issues using many predictors like this, so lets use methods to select the
beast features for our model.

# LASSO Logistic Regression for Feature Selection

```{r, echo = FALSE}
library(glmnet)

# response
y <- model_data$home_win

# predictor matrix (remove response column)
X <- model_data %>%
  select(-home_win) %>%
  as.matrix()


set.seed(42)

cv_lasso <- cv.glmnet(
  X, y,
  family = "binomial",
  alpha = 1,        # alpha = 1 → LASSO
  nfolds = 10
)

cv_lasso$lambda.min
coef(cv_lasso, s = "lambda.1se")

final_model <- glm(
  home_win ~ d_epa_off_cum + d_win_pct,
  data = train_data,
  family = binomial
)

summary(final_model)
exp(coef(final_model))

test_data$pred_prob_simple <- predict(
  final_model,
  newdata = test_data,
  type = "response"
)

test_data$pred_win_simple <- ifelse(test_data$pred_prob_simple > 0.515, 1, 0)

mean(test_data$pred_win_simple == test_data$home_win)
```

```{r}
library(pROC)

# Reduced model ROC
roc_simple <- roc(test_data$home_win, test_data$pred_prob_simple)
auc(roc_simple)
plot(roc_simple)

coords(roc_simple, "best", best.method = "youden")
```

# Results

Even though many individual predictors were not statistically significant, the 
model still predicts well because the predictors collectively capture team 
strength and matchup edges. Based on our LASSO method choosing difference in EPA
and difference in win percentage we determine the others to be redundant because
the simpler model will return similar accuracy with a simpler model.

## Limitations

Using only 2024 season limits the sample size and can make coefficients unstable.
Also some improvements could be made on the model that can also be key to a team
winning or not, such as key injuries, weather, mathcup histtory, and team 
strength, such as if team A has a very poor run defense but better record, and 
team B has a very strong run offense but a worse record, this could predict team 
A would have a good chance of winning but this would be a very key stat to the 
game that my model is missing. Things likeweather and past injury reports as 
well as team strength are very hard to obtainfor past games but in a larger 
project including things like these could add keyfeatures to a win predicting 
model.

